---
title: "Maps Tutorial"
author: "Olivia Bryant"
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Maps Tutorial

For this tutorial, we need to use several different packages.

To install each of the packages click Tools > Install Packages. You only have to install the package once.
```{r echo=FALSE, include=FALSE}
# import necessarily libraries
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(ggtext)
library(scales)
library(sf)
library(maps)
library(mapproj)
```

### Plotting Points on a Map

In order to plot the locations of the homeless GPs on the map, you'll need to create a CSV file with the latitude and longitude coordinates in. You can use get the coordinates by searching each of the GP surgeries and then right-clicking on the point on the map.

For this tutorial, we will be plotting the locations of English cathedrals on a map.

Firstly, we want to load the data:

```{r}
cathedrals <- read.csv("english_cathedrals_latlong.csv")
cathedrals
```
Let's start by plotting a map of the UK.

```{r}
# filters the map object to just look at Great Britain
UK <- map_data(map = "world", region = "UK") %>%
  filter(subregion == "Great Britain")

ggplot(data = UK, aes(x = long, y = lat, group = group)) + 
  geom_polygon() +
  coord_map()
```

```{r}
cathedrals_map <- ggplot(data = UK, aes(x = long, y = lat)) + 
  geom_polygon() +
  coord_map() +
  geom_point(data = cathedrals, 
             aes(x = LONG, 
                 y = LAT, 
                 color = "pink"))

cathedrals_map
```
We probably don't want to show the coordinates or the legend and as with any figure, we should add a title and the source.
```{r}
cathedrals_map <- ggplot(data = UK, aes(x = long, y = lat)) + 
  geom_polygon() +
  coord_map() +
  geom_point(data = cathedrals, 
             aes(x = LONG, 
                 y = LAT, 
                 color = "pink")) +
  theme(legend.position="none",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  labs(x = "",
       y = "",
       title = "Locations of English Cathedrals",
       caption= "source: ONS")

cathedrals_map
```


For more information on how to adjust the styles of R plots in ggplot, look at https://ggplot2.tidyverse.org/reference/theme.html

Finally, we can save the plot in our desired size.
```{r}
ggsave(plot = cathedrals_map, 'cathedrals_map.png', width = 5, height = 10, units = "in")
```

### Choropleth Maps

Choropleth maps are heatmaps where regions are shaded to represent data. They involve two datasets:

1. the data you wish the choropleth to show
2. files that gives the boundaries for the regions. These are typically in the form of shapefiles.


Firstly, let's read in the sample data.

```{r}
GP_sample_data <- read.csv("choropleth_sample_data.csv")
GP_sample_data
```
```{r}
NHS_regions_bound <- read_sf("Boundary Files/NHS_England_Regions_(April_2019)_EN_BFC.shp")
NHS_regions_bound
```
```{r}
regions_practice_code <- read.csv("practice_code_regions.csv")
regions_practice_code
```
We want to assign the correct region to the GP sample data and to do this, we join. We're going to use a left join (you can read more about joins here https://dplyr.tidyverse.org/reference/mutate-joins.html)
. We are joining on the practice code.

```{r}
plot_data <- left_join(GP_sample_data, regions_practice_code, by="PRACTICE_CODE")
plot_data
```

We need to find the average for each of the regions for the chloropleth. We group by the region name and then calculate the mean.

```{r}
plot_data <- plot_data %>%
  group_by(REGIONAL_OFFICE_NAME) %>%
  summarise(average_rate = mean(RATE_OVERALL))
plot_data
```

We will then join the shapefile and the data together so that the dataset contains the data we are plotting and the regions from the shapefile.
```{r}
plot_data <- left_join(NHS_regions_bound, plot_data, by = c("nhser19nm" = "REGIONAL_OFFICE_NAME"))
plot_data
```

```{r}
choropleth <- ggplot(plot_data, aes(fill = average_rate)) +
    geom_sf(color = "#ffffff", size = 0.1) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.text.y=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks=element_blank(),
    )
choropleth

```


Using the scale_fill_distiller function, you can choose a colour scheme and manually set the breaks that are listed on the legend, as well as giving the legend a title. Direction switches the direction so that the darker colours represent higher numbers.


```{r}
choropleth <- ggplot(plot_data, aes(fill = average_rate)) +
  # creates a white border around the regions with a thickness of 0.1
    geom_sf(color = "#ffffff", size = 0.1) +
    scale_fill_distiller(name="Items per 1000 Patients", palette = "Purples", direction = -1, breaks = c(0.3, 0.4, 0.5)) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.text.y=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks=element_blank(),
    )
choropleth
```
Final plot with styling:
```{r}
choropleth2 <- ggplot(plot_data, aes(fill = average_rate)) +
  # creates a white border around the regions with a thickness of 0.1
    geom_sf(color = "#ffffff", size = 0.1) +
    scale_fill_distiller(name="Items per 1000 Patients", palette = "Blues", direction = -1, breaks = c(0.3, 0.4, 0.5)) +
    theme_fivethirtyeight() +
    theme(legend.position = "right",
        legend.justification = c("right", "top"),
        legend.text = element_text(size = 8),
        plot.title = element_text(size=14),
        axis.text=element_blank(),
        axis.title = element_blank(),
        legend.direction = "vertical",
        plot.caption = element_text(hjust = 0, vjust = 0),
        plot.background = element_rect(fill = "white", colour = "white"),
        panel.background = element_rect(fill = "white", colour = "white"),
        legend.background = element_rect(fill = "white", colour = "white"),
        panel.grid.major.x = element_blank()) +
  labs(title = "Drug Prescriptions by NHS Region in England in July 2020",
       subtitle = 'The South West and Midlands show the highest prescribing rates.' ,
       x = element_blank(),
       caption = "Source: OpenPrescribing.net, EBM DataLab, University of Oxford, 2020")

choropleth2
```
```{r}
ggsave(plot = choropleth2, 'prescriptions_choropleth.png', width = 8, height = 10, units = "in")
```


### Important Design Considerations
1. Colour schemes are best when they feature one colour and have different shades of that colour, or are a gradient between two coolours. Beware of 'rainbow' colour schemes that are hard to interpret.
2. Research has shown that when looking at a plot, the human eye is drawn to text first. Always include a title, axis labels, and a caption with the data source. If you want to make a specific comment about the plot, you could add this as a subtitle.
3. When choosing sizing, save the image in your desired first. Make sure the text on the plot is no smaller than the text in the document that the figure is going into.
4. You can make any kind of styling adjustments to the plot as you want and all of these are available through a quick Google search. As a starting point, you can use the built in ggthemes package to select a pre-built theme.

### Maps to Create
1. Show the homeless GP surgeries as points on the map.
2. Produce choropleth maps for the overall prescribing rates by region and the regional differences in the subgroups. (NOTE: as you have calculated the prescribing rates of homeless and non-homeless GP separately, you will have to combine these into one dataset.)

Remember: save each sheet to a CSV to load into R.